# Activity 1: Data visualization {#sec-data-visualization}

```{r}
#| echo: false

source("_common.R")
```

## Goals

The goal of this activity is to build practice visualizing data already imported into R, and how to interpret these visualizations. 

## Set up 

**Step 1** For set up, run the following code in an R script (to open an R script, go to `File`, `New File` and click `R Script`. We are calling (as in activating) our library `tidyverse` which has some functions we will need that are not included in base R. Next, I have given you code to simulate a hypothetical data set of test scores of two schools called School A and School B. The data contains 100 scores from each school.  

```{r, message=FALSE}
library(tidyverse)


school_a = data.frame(
  score = rnorm(n = 100, mean = 80, sd = 20)
) %>% 
  mutate(school = "school_a")

school_b = data.frame(
  score = rnorm(n = 100, mean = 50, sd = 20)
) %>% 
  mutate(school = "school_b")

school_data = rbind(school_a, school_b)

```

## 1: Visualizing data: Which school teaches French better?

We can change things, though. In the code below, replace `geom_boxplot()` with `geom_point()`. Now we have points instead of a boxplot! Try `geom_jitter()`.

```{r}
school_data %>% 
  ggplot(aes(x = school, y = score)) + geom_boxplot()
```

What if we want colors? I can add `color = school`. 

```{r}
school_data %>% 
  ggplot(aes(x = school, y = score, color = school)) + geom_jitter()
```

Now, change `color` to `shape`.

```{r}
school_data %>% 
  ggplot(aes(x = school, y = score, shape = school)) + geom_jitter()
```


**Problems**:

1. Given `school_data`, how can you change both the color and shape on the basis of school? 

<p>
  <button class="btn btn-primary" type="button" data-bs-toggle="collapse" data-bs-target="#collapseWidthExample" aria-expanded="false" aria-controls="collapseWidthExample">
    Answer to question 1
  </button>
</p>
<div style="min-height: 120px;">
  <div class="collapse collapse-horizontal" id="collapseWidthExample">
  
You add an argument for both `shape` and `color` under `aes`.  
```{r}
school_data %>% 
  ggplot(aes(x = school, y = score, shape = school, color = school)) + geom_jitter()
```

  </div>
</div>

2. Given `school_data`, how can you change the color on the basis of score, rather than school?

<p>
  <button class="btn btn-primary" type="button" data-bs-toggle="collapse" data-bs-target="#collapseWidthExample2" aria-expanded="false" aria-controls="collapseWidthExample2">
    Answer to question 2
  </button>
</p>
<div style="min-height: 120px;">
  <div class="collapse collapse-horizontal" id="collapseWidthExample2">
  
You change `color` to `score` in the `aes` layer. 
```{r}
school_data %>% 
  ggplot(aes(x = school, y = score, color = score)) + geom_jitter()
```

  </div>
</div>

3. Given `school_data`, create a box plot which in which the data are colored according to school.  

<p>
  <button class="btn btn-primary" type="button" data-bs-toggle="collapse" data-bs-target="#collapseWidthExample3" aria-expanded="false" aria-controls="collapseWidthExample3">
    Answer to question 3
  </button>
</p>
<div style="min-height: 120px;">
  <div class="collapse collapse-horizontal" id="collapseWidthExample3">
  
```{r}
school_data %>% 
  ggplot(aes(x = school, y = score, color = school)) + geom_boxplot()
```

  </div>
</div>

