# Final Project: Statistics for Linguistcs 
## SoSE 2024

The general due date for this activity is **July 24th**, but if you need more time please discuss this with me. 
To turn in your work, please save your R script and title it using your last name (e.g. Parrish.R), and also include a document (word, pdf, Google doc etc.) which includes a description of the data you chose, your visualization and your reporting of the descriptive and inferential statistics. 

Send these two files to me via email: `Parrish@em.uni-frankfurt.de`

Here is an [example](https://kparrish92.github.io/stats/05_example.html) in HTML format.

```{r}
#| echo: false

source("_common.R")
```


**Previous activities for reference**

Week 1 [Data Vislualization](https://kparrish92.github.io/stats/01_data_vis.html)

Week 2 [Descriptive Statistics](https://kparrish92.github.io/stats/02_desc_stats.html)

Week 3 [T-tests](https://kparrish92.github.io/stats/03_t_test.html)

## Goals

The goal of this activity is to review and better commit to memory lessons 1-3 (data visualization, descriptive statistics and the t-test). Using openly available data, you will 1) create a visualiztion, 2) report some descriptive statistics and 3) run a t-test and report the results.
Rather than a full demonstration this time, I'll give you hints which functions to use and get you started when necessary.

## Set up 

Set up is the same as always: Open an R Script (`File`, `New File` click `R Script`). For set up, run the following code in the newly opened R script to load the tidyverse and untidydata packages.

```{r, message=FALSE}
library(tidyverse)
library(untidydata)
```

The untidydata library is a collection of 5 linguistic data sets. For this assignment, you will need to choose one of the five (details on the data sets: https://www.jvcasillas.com/untidydata/).

## Example: The VOT data set

If you simply type "untidydata::" in an R script, you should see a drop down window with the names of the 5 data sets. 
You can simply click on one or press the down arrow key followed by enter to select one. 
It is a good idea to save this data to a new object, just so it shows up in your global environment. 
I'll call mine `vot_data`.
In sum, you can just type `untidydata::` and one of the 5 data set names to print the data. 
Set what ever you choose equal to a variable name and run it to save it in the global environment. 

```{r}
vot_dat = untidydata::vot
```

## 1. Create a visualization 

You will want to use the `ggplot` function. Think about what kind of visualizations are possible for the data you chose. Many plots generated by ggplot2 require you to define an x and y axis. You could choose two continous variables and create a scatter plot using `geom_point()`, or if one variable if continous and the other is categorical, `geom_boxplot()`. 

GGplot cheatsheet: https://www.maths.usyd.edu.au/u/UG/SM/STAT3022/r/current/Misc/data-visualization-2.1.pdf 


## 2. Report some descriptive statistics (mean and sd)
 
You need to use the functions `mean()` and `sd()`. 
It might be that your data has multiple conditons which are reported in one column. If this is the case, you have two options. You can use the `group_by()` and `summarize()` functions, or you can filter your data using the `filter()` function. 

## 3. Run and report a t-test 

Of course, you'll use the `t.test()` function. You will need to choose continuous data that you can somehow separate. You can filter the bigger data set here. 