# Final Project: Statistics for Linguistcs 
## SoSE 2025

Report an analysis of data with visualizations (ggplot), descriptive statistics (means and standard deviations), and an inferential test (a t-test or a linear model).
The general due date for this activity is **August 22nd**, but if you need more time please discuss this with me. 
To turn in your work, please save your R script and title it using your last name (e.g. Parrish.R), and also include a document (word, pdf, Google doc etc.) which includes a description of the data you chose, your visualization and your reporting of the descriptive and inferential statistics. 

Send these two files to me via email: `Parrish@em.uni-frankfurt.de`

Here is an [example](https://kparrish92.github.io/stats/05_example.html) in HTML format.

```{r}
#| echo: false

source("_common.R")
```


**Previous activities for reference**

Week 1 [Data Vislualization](https://kparrish92.github.io/stats/01_data_vis.html)

Week 2 [Descriptive Statistics](https://kparrish92.github.io/stats/02_desc_stats.html)

Week 3 [T-tests](https://kparrish92.github.io/stats/03_t_test.html)

## Goals

The goal of this activity is to review and better commit to memory lessons 1-3 (data visualization, descriptive statistics and the t-test). Using openly available data, you will 1) create a visualization, 2) report some descriptive statistics and 3) run a t-test and report the results.
Rather than a full demonstration this time, I'll give you hints which functions to use and get you started when necessary.

## Set up 

Set up is the same as always: Open an R Script (`File`, `New File` click `R Script`). For set up, run the following code in the newly opened R script to load the tidyverse and untidydata packages.

```{r, message=FALSE}
library(tidyverse)
library(untidydata)
```

The untidydata library is a collection of 5 linguistic data sets. For this assignment, you will need to choose one of the five (details on the data sets: https://www.jvcasillas.com/untidydata/).

OR you can find an open data set to use: https://osf.io/ 

## Example: The VOT data set

If you simply type "untidydata::" in an R script, you should see a drop down window with the names of the 5 data sets. 
You can simply click on one or press the down arrow key followed by enter to select one. 
It is a good idea to save this data to a new object, just so it shows up in your global environment. 
I'll call mine `vot_data`.
In sum, you can just type `untidydata::` and one of the 5 data set names to print the data. 
Set what ever you choose equal to a variable name and run it to save it in the global environment. 

```{r}
vot_dat = untidydata::vot
```

## 1. Create a visualization 

You will want to use the `ggplot` function. Think about what kind of visualizations are possible for the data you chose. Many plots generated by ggplot2 require you to define an x and y axis. You could choose two continuous variables and create a scatter plot using `geom_point()`, or if one variable if continuous and the other is categorical, `geom_boxplot()`. 

GGplot cheatsheet: https://www.maths.usyd.edu.au/u/UG/SM/STAT3022/r/current/Misc/data-visualization-2.1.pdf 


## 2. Report some descriptive statistics (mean and sd)
 
You need to use the functions `mean()` and `sd()`. 
It might be that your data has multiple conditions which are reported in one column. If this is the case, you have two options. You can use the `group_by()` and `summarize()` functions, or you can filter your data using the `filter()` function. 

## 3. Run and report a t-test 

Of course, you'll use the `t.test()` function. You will need to choose continuous data that you can somehow separate. You can filter the bigger data set here. 

## Linear Model: Extra credit & challenge 

If you want a challenge, find an open data set and run a linear model using the `lm()` function in R. 
You should describe the model structure. What is the outcome variable? (What is being predicted), and what are the predictor variables?
The model tests the relationship between them.
What's the intercept of the model represent? What does the other (second) term represent and how do we know it's meaningful (significant)?
For extra extra credit - can we add more than one variable as a predictor? 

