---
title: "Week 3: Data Visualization"
subtitle: 'Statistics for Linguists'
author  : "Kyle Parrish"
date    : "Goethe Uni Frankurt</br>Summer 2025</br>Last update: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, hygge, my-css.css, rutgers-fonts]
    lib_dir: libs
    nature:
      countIncrementalSlides: false
      ratio: 16:9
---

```{r, include = FALSE}
library(here)
library(tidyverse)
```


```{r xaringanExtra-clipboard, echo=FALSE}
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

# Data visualization

--

.large[
.pull-left[
Data visualization is a vital step in any statistical analysis.

**Visualization can give us important insights regarding relationships**
]]

--

.large[
.pull-right[
**Ggplot 2**: The R package used in most plots. See the ggplot cheatsheet [here](https://rstudio.github.io/cheatsheets/data-visualization.pdf) 

Different plot types are ideal for categorical and continuous variables.
]]

---

class: title-slide-section, middle

## Part 1: Basics 

---

## Part 1: Basics 


.large[
Ggplot is a function in R that requires for data, mapping (coordinates), and a plot type.
]

--

.large[
We will use `mtcars` as an example data set. It is a built in data set in R that contains many different variable types.
]


```{r}
glimpse(mtcars)
```

---

class: title-slide-section, middle

## Part 2: One variable  

---

## Part 2: One variable  

.pull-left[
**A Histogram** is often use to visualize a distribution. 

This plot shows the relative quantity of data points between a certain interval (bin) 

We plot the variable *cyl* here, which only has 4, 6 or 8.
]


--

.pull-right[
```{r, out.width = '400px', message = FALSE}
# Histogram
mtcars %>% 
  ggplot(aes(x = cyl)) + geom_histogram()
```
]

---


## Part 2: One variable  

.pull-left[
**A dotplot** also plots a one variable distribution.

This plot shows each the same information as the histogram, but with clearer distinctions of individual data points (better for small data sets).
]

--

.pull-right[
```{r, out.width = '400px', message = FALSE}
# Histogram
mtcars %>% 
  ggplot(aes(x = cyl)) + geom_dotplot()
```
]


---

## Part 3: Two variables (one categorical one continuous)

.pull-left[
**A barplot** compares quantities of two categories.
]

--

.pull-right[
```{r, out.width = '400px', message = FALSE}
# Histogram
mtcars %>% 
  group_by(cyl) %>% 
  summarise(mean_mpg = mean(mpg)) %>% 
  ggplot(aes(y = mean_mpg, x = as.factor(cyl))) + geom_col()
```
]

---

## Part 3: Two variables (one categorical one continuous)

.pull-left[
**A boxplot** shows the distribution of values in various categories. The box itself contains 50% of the data points while the median is line within the box.
]

--

.pull-right[
```{r, out.width = '400px', message = FALSE}
# Histogram
mtcars %>% 
  ggplot(aes(y = mpg, x = as.factor(cyl))) + geom_boxplot()
```
]

---

## Part 4: Two variables (one categorical, one continuous)

.pull-left[
**A scatter plot** shows the relationship between two continuous variables.

Here we plot the relationship between mpg and weight.
]

--

.pull-right[
```{r, out.width = '400px', message = FALSE}
# Histogram
mtcars %>% 
  ggplot(aes(y = mpg, x = wt)) + geom_point()
```
]

---

## Part 4: Two variables (one categorical, one continuous)

.pull-left[
**A line plot** also shows the relationship between two continuous variables.

We plot a line of best fit to the data points.
]

--

.pull-right[
```{r, out.width = '400px', message = FALSE}
# Histogram
mtcars %>% 
  ggplot(aes(y = mpg, x = wt)) + geom_smooth(method = "lm")
```
]

---


## Part 4: Two variables (one categorical, one continuous)

.pull-left[
Ggplots also have layers - we can add both the points and line of best fit to the plot.
]

--

.pull-right[
```{r, out.width = '400px', message = FALSE}
# Histogram
mtcars %>% 
  ggplot(aes(y = mpg, x = wt)) + geom_point() + geom_smooth(method = "lm")
```
]

---

## Part 4: Three variables (Two categorical, one continuous)

.pull-left[
We can also separate categories within the previous plot - let's say we want to look at whether the relationship between weight and mpg is similar for different cylinder types.
]

--

.pull-right[
```{r, out.width = '400px', message = FALSE, warning=FALSE}
# Histogram
mtcars %>% 
  ggplot(aes(y = mpg, x = wt, color = as.factor(cyl))) + geom_point() + geom_smooth(method = "lm")
```
]