---
title: "Week 1: Distributions, means and standard deviations"
subtitle: 'Statistics for Linguists'
author  : "Kyle Parrish"
date    : "Goethe Uni Frankurt</br>Summer 2025</br>Last update: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, hygge, my-css.css, rutgers-fonts]
    lib_dir: libs
    nature:
      countIncrementalSlides: false
      ratio: 16:9
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '', fig.width = 6, fig.height = 6, message = FALSE, warning = FALSE, echo = FALSE
)
```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_duo_accent(
  # primary_color = "#345865",
  primary_color = "#B56B6F",
  secondary_color = "#2c8475",
  black_color = "#4242424",
  white_color = "#FFF",
  base_font_size = "25px",
  # text_font_family = "Jost",
  # text_font_url = "https://indestructibletype.com/fonts/Jost.css",
  header_font_google = google_font("Times New Roman", "400", "700"),
  header_font_weight = "400",
  inverse_header_color = "#eaeaea",
  title_slide_text_color = "black",
  text_slide_number_color = "#9a9a9a",
  text_bold_color = "#f79334",
  code_inline_color = "#B56B6F",
  code_highlight_color = "transparent",
  link_color = "#2c8475",
  table_row_even_background_color = lighten_color("#345865", 0.9),
  extra_fonts = list(
    google_font("Times New Roman", "400")
  ),
  colors = c(
    green = "#31b09e",
    "green-dark" = "#2c8475",
    highlight = "#87f9bb",
    purple = "#887ba3",
    pink = "#B56B6F",
    orange = "#f79334",
    red = "#dc322f",
    `blue-dark` = "#002b36",
    `text-dark` = "#202020",
    `text-darkish` = "#424242",
    `text-mild` = "#606060",
    `text-light` = "#9a9a9a",
    `text-lightest` = "#eaeaea"
  ),
  extra_css = list(
    ".remark-slide-content h3" = list(
      "margin-bottom" = 0, 
      "margin-top" = 0
    ),
    ".smallish, .smallish .remark-code-line" = list(`font-size` = "0.9em")
  )
)
xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "share_again"))
xaringanExtra::use_extra_styles()
```


```{r components, include=FALSE}
slides_from_images <- function(
  path,
  regexp = NULL,
  class = "hide-count",
  background_size = "contain",
  background_position = "top left"
) {
  if (isTRUE(getOption("slide_image_placeholder", FALSE))) {
    return(glue::glue("Slides to be generated from [{path}]({path})"))
  }
  if (fs::is_dir(path)) {
    imgs <- fs::dir_ls(path, regexp = regexp, type = "file", recurse = FALSE)
  } else if (all(fs::is_file(path) && fs::file_exists(path))) {
    imgs <- path
  } else {
    stop("path must be a directory or a vector of images")
  }
  imgs <- fs::path_rel(imgs, ".")
  breaks <- rep("\n---\n", length(imgs))
  breaks[length(breaks)] <- ""

  txt <- glue::glue("
  class: {class}
  background-image: url('{imgs}')
  background-size: {background_size}
  background-position: {background_position}
  {breaks}
  ")

  paste(txt, sep = "", collapse = "")
}
options("slide_image_placeholder" = FALSE)
```

# Warm up: 10 minutes

.pull-left[
.full-width[
.content-box-grey[
Without looking, what would you guess are the differences between **descriptive** and **inferential** statistics?

Why do we need both?
]]]

---

# Warm up

.pull-left[
.full-width[
.content-box-grey[
**Descriptive Statistics** are a summary of the actual collected data sample (such as the mean and standard deviation). These are directly measured.
]]]

--

.pull-right[
.full-width[
.content-box-grey[
**Inferential Statistics** test hypotheses on the basis of a sample of data and how much information can be inferred by the data (e.g. is there a difference between two points of comparison)?
]]]

---

# Example: Test scores 

.large[
We want to know if .blue[school A] has better test scores than .red[school B]
]

--

.large[
We ask 10 people their test scores at both schools:
]

```{r, echo=TRUE}
# Scores range from 0-100 in percentage correct
scores_school_a = c(65, 42, 91, 95, 71, 54, 90, 82, 73, 80)

scores_school_b = c(71, 98, 82, 51, 44, 68, 80, 70, 81, 63)
```
--

```{r, echo = TRUE}
# means of the schools 
mean(scores_school_a)
mean(scores_school_b)
```

---

# Example: Test scores 

.large[
The descriptive statistics suggest that .blue[school A] does indeed have better test scores than .red[school B]...but can we trust this difference?
]

--

.large[
In this case, we have a **mean difference** of 
`r round(mean(scores_school_a) - mean(scores_school_b), digits = 2)`]

--

.large[
Importantly, a given mean difference has much more meaning in the context of its variation: one measure of variation is the **standard deviation**.
]

---

# Standard Deviation

.large[
For normally distributed data, 68% of the data points fall within +/- 1 standard deviation of the mean.
]

--

```{r, echo = TRUE}
# Standard deviations of both schools 
sd(scores_school_a)
sd(scores_school_b)
```


---

background-image: url(./images/school_a_low.svg)
background-size: contain

---

background-image: url(./images/school_b_low.svg)
background-size: contain

---

background-image: url(./images/school_both_low.svg)
background-size: contain

---


# Is School A better?

.large[
We have a small sample that make it difficult to tell whether the measured difference is reliable or due to random error (statistical power).
]

--

.large[
Let's say we try the experiment again, but this time we get 100 tests:
]


---

background-image: url(./images/school_both_high.svg)
background-size: contain

---

background-image: url(./images/school_both_high_eff.svg)
background-size: contain

