---
title: "Week 5: The Linear Model"
subtitle: 'Statistics for Linguists'
author  : "Kyle Parrish"
date    : "Goethe Uni Frankurt</br>Summer 2025</br>Last update: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, hygge, my-css.css, rutgers-fonts]
    lib_dir: libs
    nature:
      countIncrementalSlides: false
      ratio: 16:9
---

```{r, include = FALSE}
# Packages --------------------------------------------------------------------

# Custom
library("ds4ling")
library("lingStuff")
library("untidydata")
library("academicWriteR")

# Tidyverse
library("dplyr")
library("tidyr")
library("readr")
library("purrr")
library("here")
library("forcats")
library("stringr")
library("glue")

# Modeling
library("lme4")
library("brms")

# Plotting
library("ggplot2")
library("patchwork")
library("ggridges")
library("ggrepel")
library("viridis")
library("ggfortify")

# Misc
library("knitr")
library("kableExtra")
library("latex2exp")
library("broom")
library("fGarch")
library("equatiomatic")
library("countdown")
library("tiktokrmd")
```


```{r xaringanExtra-clipboard, echo=FALSE}
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r}
#| echo: false
source("_common.R")
```


<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
.center[
## Slides are created from open course notes by Joseph Casillas [here](https://www.ds4ling.jvcasillas.com/slides/05_lm/01_bivariate_regression/#/title-slide/=)]

---

# Linear Regression

```{r include=FALSE}
knitr::opts_chunk$set(cache=F, warning=FALSE, message=FALSE, echo = FALSE)
library(arm)
library(ggplot2)
library(tidyr)
library(dplyr)
options(scipen = 0)
```

.pull-left[
.large[
So far, we have been focusing on **comparing two means** (e.g. is class A better than class B at math).
]
]

--

.pull-right[
.large[
But what about .red[confounding variables?]

What if the differences in math scores between two schools are not simply because one school is better, but because of another variable (socio-economic status)?

For this we can use a linear model - specifically a multivariate linear regression. 

It can consider several variables at once! 
]]

---

# What are models?  

.large[
> "All models are wrong, but some are useful." <br><br>
George Box

]

.large[
**Models** simplify reality for purposes of understanding or prediction.  
]

--

.large[

Statistical modeling has these two sometimes divergent goals:

- .green[*Description*]: using a model to describe the relationship between an outcome variable of interest and one or more predictor variables.


- .red[*Prediction*]: using a model to predict unknown instances of the outcome variable such that out-of-sample predictive error is minimized.
]

---

# The basics

.pull-left[
.large[
A linear model is *parametric* because we assume that the relationship between two variables is linear and can be defined by the *parameters* of a line---an .green[intercept] and .red[slope].  We'll start by considering a simple linear model.
]]

--

.pull-right[
```{r}
mtcars %>% 
ggplot(aes(x = mpg, y = wt)) + geom_point() + geom_smooth(method = "lm")
```
]

---

# Simple linear model

.large[
A simple linear model has an outcome, $y$, and one predictor, $x$. 

$$y = mx + b$$
]


--

We often want to *predict* or *describe* the outcome .blue[y] as a function of the predictor .red[x].

--

.large[
**Examples**

.blue[weight] ~ .red[height]

.blue[IQ] ~ .red[age]

.blue[ice cream sold] ~ .red[temperature]

.blue[RT] ~ .red[group]
]

---

# Linear algebra 

*mx* is the *slope* and .blue[b] is the point where the line crosses the y-axis when x = 0, i.e. the *y-intercept*.

If you know two of the variables, you can solve for the third... 
assume x = 2. Solve for y.

--

$$y = 50 + 10x$$

--

$$y = 50 + 10 \times 2$$

--

$$y = 50 + 20$$

--

$$y = 70$$

---

# Cartesian coordinates

.pull-left[
```{r, echo=FALSE}
#| label: cartesian_coord_plot1
#| fig-asp: 0.8
y_hashes <- data.frame(
  x = -0.15, 
  x_end = 0.15, 
  y_axis = seq(-5, 5, 1), 
  label = 'y-axis'
)

x_hashes <- data.frame(
  y = -0.15, 
  y_end = 0.15, 
  x_axis = seq(-5, 5, 1), 
  label = 'x-axis'
)

coords_p1 <- tribble(
  ~'x', ~'y', ~'label', 
    1,    4,   'y-axis', 
    4,   -1,   'x-axis'
  ) |>
  ggplot() + 
  aes(x = x, y = y, label = label, color = label) + 
  geom_text(size = 10, show.legend = FALSE) + 
  scale_x_continuous(
    breaks = seq(-5, 5, 1), 
    labels = seq(-5, 5, 1)
  ) + 
  scale_y_continuous(
    breaks = seq(-5, 5, 1), 
    labels = seq(-5, 5, 1)
  ) +
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0) + 
  geom_segment(
    data = y_hashes, 
    aes(
      x = x, xend = x_end, 
      y = y_axis, yend = y_axis, 
      label = NULL, color = NULL
    ), 
    show.legend = FALSE
  ) +
  geom_segment(
    data = x_hashes, 
    aes(
      x = x_axis, xend = x_axis, 
      y = y, yend = y_end, 
      label = NULL, color = NULL), 
    show.legend = FALSE
  ) +
  scale_color_brewer(palette = "Set1") + 
  ds4ling_bw_theme(base_size = 24)

coords_p1
```
]


.pull-right[
```{r, echo=FALSE}
#| label: cartesian_coord_plot2
#| fig-asp: 0.8
coords_p1 + geom_abline(
  intercept = 0, 
  slope = 1, 
  linewidth = 1.5,
  lty = 1, 
  color = "darkred"
  )
```
]
---

# Cartesian coordinates

.pull-left[
```{r, echo=FALSE}
#| label: cartesian_coord_plot3
#| fig-asp: 0.8
coords_p1 + geom_abline(
  intercept = 1, 
  slope = 3, 
  linewidth = 1.5,
  lty = 1, 
  color = "darkred"
  )
```
]

--

.pull-right[
```{r, echo=FALSE}
#| label: cartesian_coord_plot4
#| fig-asp: 0.8
coords_p1 + geom_abline(
  intercept = 2, 
  slope = -1, 
  linewidth = 1.5,
  lty = 1, 
  color = "darkred"
  )
```
]

---

# Bivariate regression

The linear model is basically the same as linear algebra, with two subtle differences
  1. We fit the line through data points (measurements, observations)
  2. We use slightly different terminology


$$\color{blue}{response} \sim intercept + (slope * \color{red}{predictor})$$

$$\color{blue}{\hat{y}} = a + b\color{red}{x}$$

--

- We call our dependent variable the .blue[outcome variable] (or criterion)
- Our independent variables are called .red[predictors]
- The intercept and the slope are *coefficients*
  - These are the meat and potatoes of the linear model
  - They are also called *parameter estimates*

---

# How do we know the line of best fit?

.pull-left[
```{r, echo=FALSE}
#| label: best_fit1
#| fig-asp: 0.8

coords_p2_labels <- tribble(
  ~'x', ~'y', ~'label', 
    1,    4,   'y-axis', 
    4,   -1,   'x-axis'
  )

coords_p2_points <- tribble(
  ~'x', ~'y',
    1,    1,
    2,    2, 
    3,    3
  )

coords_p2 <- ggplot(coords_p2_labels) + 
  aes(x = x, y = y, label = label) + 
  geom_text(size = 8, show.legend = FALSE) + 
  geom_point(
    data = coords_p2_points,  
    aes(x = x, y = y, label = NULL), 
    color = "blue", size = 3
  ) +
  scale_x_continuous(
    breaks = seq(-5, 5, 1), 
    labels = seq(-5, 5, 1)
  ) + 
  scale_y_continuous(
    breaks = seq(-5, 5, 1), 
    labels = seq(-5, 5, 1)
  ) +
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0) + 
  geom_segment(
    data = y_hashes, 
    aes(
      x = x, xend = x_end, 
      y = y_axis, yend = y_axis, 
      label = NULL, color = NULL), 
    show.legend = FALSE
  ) +
  geom_segment(
    data = x_hashes, 
    aes(
      x = x_axis, xend = x_axis,  
      y = y, yend = y_end, 
      label = NULL, color = NULL), 
    show.legend = FALSE
  ) + 
  scale_color_brewer(palette = "Set1") +
  ds4ling_bw_theme(base_size = 24)

coords_p2
```
]

.pull-right[
```{r, echo=FALSE}
#| label: best_fit2
#| fig-asp: 0.8
coords_p2 + geom_abline(intercept = 0, slope = 1, color = "darkred")
```
]

---

# Bivariate Regression

.pull-left[
```{r, echo=FALSE}
#| label: best_fit3
#| fig-asp: 0.8

ggplot(coords_p2_points) + 
  aes(x = x, y = y) + 
  coord_cartesian(xlim = c(0, 4), ylim = c(0, 4)) + 
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0) + 
  geom_hline(
    yintercept = mean(coords_p2_points$y), lty = 1, 
    color = 'blue', linewidth = 1.5
  ) +
  geom_abline(
    intercept = 0, slope = 1, lty = 2, 
    color = "darkred", linewidth = 0.7
  ) + 
  geom_segment(
    aes(x = 1, xend = 1, y = 1, yend = 2), 
    lty = 3, linewidth = 0.8
  ) + 
  geom_segment(
    aes(x = 3, xend = 3, y = 2, yend = 3), 
    lty = 3, linewidth = 0.8
  ) + 
  geom_point(color = 'blue', fill = 'grey70', pch = 21, size = 6) + 
  annotate("text", x = 0.5, y = 1, label = "(1, 1)", lwd = 6, 
    color = 'darkgrey') +
  annotate("text", x = 3.5, y = 3, label = "(3, 3)", lwd = 6, 
    color = 'darkgrey') +
  ds4ling_bw_theme(base_size = 24)
```
]

.pull-right[
```{r, echo=FALSE}
#| label: best_fit_table1
#| results: 'asis'
kable(coords_p2_points, format = "html")
```


- If we tried to predict using just the .blue[mean of y] we would be right once
- We'd miss badly for .red[(1, 1)] and .red[(3, 3)].
- The **line of best fit** is that which reduces the distance between the *predicted values* of `y` and the *observed values* of `y`. 
]


---

# Bivariate regression

### Measurement error

.pull-left[
- We rarely (never) find a perfectly linear relationship in our data
  - Most relationships are not perfectly linear
  - Our measurements are not perfect (VOT, formants, durations, RT)
  - There is error in everything (normal distribution)
- We account for *error* in our models]


### $$\hat{y} = a + bx + \color{red}{\epsilon}$$

---

# Bivariate regression

### Measurement error

.pull-left[
```{r, echo=FALSE}
#| label: best_fit4
#| fig-asp: 0.8
imperf_points1 <- tribble(
  ~'x', ~'y', ~'fitted1', ~'fitted2', 
    1,    2,    1.6,        0.75, 
    2,    1,    2.7,        1.25, 
    3,    3,    3.8,        1.75
)

ggplot(imperf_points1) + 
  aes(x = x, y = y) + 
  coord_cartesian(xlim = c(0, 4), ylim = c(0, 4)) + 
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0) + 
  geom_hline(
    yintercept = mean(imperf_points1$y), 
    lty = 1, color = 'blue', linewidth = 1.5, alpha = 0.3
  ) +
  geom_point(color = 'blue', fill = 'grey70', pch = 21, size = 6) + 
  ds4ling_bw_theme(base_size = 20)

```
]

.pull-right[
- Again, we can see that the .blue[mean of `y`] isn't the best solution. 
  - It can summarize the `y` data
  - It cannot explain the relationship between `x` and `y`.
- We need a method to assess how well a given line fits the data
- We need a method to determine the optimal intercept and slope (the line of best fit)
]

---

# Bivariate regression

### More on error

```{r, fig.show="hold", out.width="33%"}

imperf_plot0 <- ggplot(imperf_points1) + 
  aes(x = x, y = y) + 
  coord_cartesian(xlim = c(0, 4), ylim = c(0, 4)) + 
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0) + 
  geom_hline(
    yintercept = mean(imperf_points1$y), 
    lty = 1, color = 'blue', linewidth = 1.5, alpha = 0.1
  ) +
  ds4ling_bw_theme(base_size = 12)

imperf_plot1 <- imperf_plot0 + 
  geom_abline(
    intercept = 0.5, slope = 1.1, 
    lty = 2, color = "darkred"
  ) + 
  geom_segment(
    aes(xend = x, yend = fitted1), 
    lty = 3, linewidth = 0.6
  ) + 
  geom_point(color = 'blue', fill = 'grey70', pch = 21, size = 3)

imperf_plot2 <- imperf_plot0 + 
  geom_abline(
    intercept = 0.25, slope = 0.5, 
    lty = 2, color = "darkred"
  ) +
  geom_segment(
    aes(xend = x, yend = fitted2), 
    lty = 3, linewidth = 0.6
  ) + 
  geom_point(color = 'blue', fill = 'grey70', pch = 21, size = 3)

imperf_mod <- lm(y ~ x, data = imperf_points1)

imperf_plot3 <- imperf_plot0 + 
  geom_abline(
    intercept = coef(imperf_mod)[1], 
    slope = coef(imperf_mod)[2], 
    lty = 2, color = "darkred"
  ) + 
  geom_segment(
    aes(xend = x, yend = fitted(imperf_mod)), 
    lty = 3, linewidth = 0.6
  ) + 
  geom_point(color = 'blue', fill = 'grey70', pch = 21, size = 3)

imperf_plot1
imperf_plot2
imperf_plot3
```

- One way to do this is to calculate the 'distance' between predicted y (ŷ, y-hat) and observed y ($y_i$). 
- If we add up this distance for each observation we can compare it to other lines (i.e., other distances). 
- The line that produces the shortest distance is the best. 


---

# Bivariate regression

### More on error

.pull-left[
```{r, echo=FALSE}
#| label: best_fit6
#| fig-asp: 0.9
imperf_plot1a <- imperf_plot1 + 
  geom_point(color = 'blue', fill = 'grey70', pch = 21, size = 6) + 
  ds4ling_bw_theme(base_size = 26)
imperf_plot1a
```
]

.pull-right[
- The difference between $\hat{y}$ and $y_i$ is called *prediction error*
- The *total prediction error* (TPE) is the sum of the prediction error of all observations
- This measurement is not ideal because negative values cancel out the positive ones, thus we square them. 
- We call this the sum of square of the errors (SSE)
]

---

# Bivariate regression

### More on error

.pull-left[
```{r}
#| label: best_fit7
#| fig-asp: 0.9
imperf_plot1a + 
  annotate(
    geom = "text", 
    x = 0.85, y = 1.85, 
    label = "[", size = 18, 
    color = 'grey70'
  ) +
  geom_segment(aes(x = 2.17, xend = 2.17, y = 1, yend = 2.7), 
    linewidth = 1.7, color = 'grey70') +
  geom_segment(aes(x = 2.1, xend = 2.19, y = 1, yend = 1), 
    linewidth = 1.7, color = 'grey70') +
  geom_segment(aes(x = 2.1, xend = 2.19, y = 2.7, yend = 2.7), 
    linewidth = 1.7, color = 'grey70') +
  geom_segment(aes(x = 3.17, xend = 3.17, y = 3, yend = 3.8), 
    linewidth = 1.7, color = 'grey70') + 
  geom_segment(aes(x = 3.1, xend = 3.19, y = 3, yend = 3), 
    linewidth = 1.7, color = 'grey70') +
  geom_segment(aes(x = 3.1, xend = 3.19, y = 3.8, yend = 3.8), 
    linewidth = 1.7, color = 'grey70') 
```
]

.pull-right[
```{r, fig.show="hold", out.width="33%"}
imperf_table1 <- imperf_points1 |> 
  dplyr::select(x_i = x, y_i = y, y_hat = fitted1) |> 
  mutate(pred.error = y_i - y_hat, sqrd.error = pred.error ^ 2)

pred_error1 <- sum(imperf_table1$pred.error)
sqrd_error1 <- sum(imperf_table1$sqrd.error)

imperf_table1 |> 
  add_row(y_hat = NA) |> 
  add_row(
    y_hat = NA, 
    pred.error = pred_error1, 
    sqrd.error = sqrd_error1
  ) |>
  mutate(
    x_i = cell_spec(
      x = x_i, 
      format = "html", 
      color = if_else(is.na(x_i), "white", "black")
    ), 
    y_i = cell_spec(
      x = y_i, 
      format = "html", 
      color = if_else(is.na(y_i), "white", "black")
    ), 
    y_hat = cell_spec(
      x = y_hat, 
      format = "html", 
      color = if_else(is.na(y_hat), "white", "black")
    )
  ) |> 
  kable(format = "html", escape = F) |> 
  row_spec(4, color = "white") |> 
  row_spec(5, color = "#cc0033")
```

This line is not the best fit for these data because it does not *minimize the sum of the squares of the errors*.
]

---

# Bivariate regression

### More on error

.pull-left[
```{r}
#| label: best_fit8
#| fig-asp: 0.9
imperf_plot2a <- imperf_plot2 + 
  geom_point(color = 'blue', fill = 'grey70', pch = 21, size = 6) + 
  ds4ling_bw_theme(base_size = 26)
imperf_plot2a
```
]


.pull-right[
```{r}
#| label: best_fit_table3
#| results: 'asis'
imperf_table2 <- imperf_points1 |> 
  dplyr::select(x_i = x, y_i = y, y_hat = fitted2) |> 
  mutate(pred.error = y_i - y_hat, sqrd.error = pred.error ^ 2)

pred_error2 <- sum(imperf_table2$pred.error)
sqrd_error2 <- sum(imperf_table2$sqrd.error)

imperf_table2 |> 
  add_row(y_hat = NA) |> 
  add_row(
    y_hat = NA, 
    pred.error = pred_error2, 
    sqrd.error = sqrd_error2
  ) |>
  mutate(
    x_i = cell_spec(
      x = x_i, 
      format = "html", 
      color = if_else(is.na(x_i), "white", "black")
    ), 
    y_i = cell_spec(
      x = y_i, 
      format = "html", 
      color = if_else(is.na(y_i), "white", "black")
    ), 
    y_hat = cell_spec(
      x = y_hat, 
      format = "html", 
      color = if_else(is.na(y_hat), "white", "black")
    )
  ) |> 
  kable(format = "html", escape = F) |> 
  row_spec(4, color = "white") |> 
  row_spec(5, color = "#cc0033")
```

- This line is not the best fit for these data because it does not *minimize the 
sum of the squares of the errors*. 
]

---

# Bivariate regression

### More on error


.pull-left[
```{r}
#| label: best_fit9
#| fig-asp: 0.9
imperf_mod <- lm(y ~ x, data = imperf_points1)

imperf_plot_bestfit <- ggplot(imperf_points1) + 
  aes(x = x, y = y) + 
  coord_cartesian(xlim = c(0, 4), ylim = c(0, 4)) + 
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0) + 
  geom_hline(
    yintercept = mean(imperf_points1$y), 
    lty = 1, color = 'blue', 
    linewidth = 1.5, alpha = 0.1
  ) +
  geom_abline(
    intercept = coef(imperf_mod)[1], 
    slope = coef(imperf_mod)[2], 
    lty = 2, color = "darkred"
  ) + 
  geom_segment(
    aes(xend = x, yend = fitted(imperf_mod)), 
    lty = 3, linewidth = 1
  ) +
  geom_point(color = 'blue', fill = 'grey70', pch = 21, size = 6) + 
  ds4ling_bw_theme(base_size = 26)
imperf_plot_bestfit
```
]

.pull-right[
```{r}
#| label: best_fit_table4
#| results: 'asis'
imperf_table3 <- imperf_points1 |> 
  dplyr::select(x_i = x, y_i = y) |>
  mutate(
    y_hat = fitted(imperf_mod), 
    pred.error = y_i - y_hat, 
    sqrd.error = pred.error^2
  )

pred_error3 <- sum(imperf_table3$pred.error)
sqrd_error3 <- sum(imperf_table3$sqrd.error)

imperf_table3 |> 
  add_row(y_hat = NA) |> 
  add_row(
    y_hat = NA, 
    pred.error = pred_error3, 
    sqrd.error = sqrd_error3
  ) |>
  mutate(
    x_i = cell_spec(
      x = x_i, 
      format = "html", 
      color = if_else(is.na(x_i), "white", "black")
    ), 
    y_i = cell_spec(
      x = y_i, 
      format = "html", 
      color = if_else(is.na(y_i), "white", "black")
    ), 
    y_hat = cell_spec(
      x = y_hat, 
      format = "html", 
      color = if_else(is.na(y_hat), "white", "black")
    )
  ) |> 
  kable(format = "html", escape = F) |> 
  row_spec(4, color = "white") |>
  row_spec(5, color = "#cc0033")
```

- This line represents a much better fit (1.50 is lower than the SSE of the other lines)
- So how do we calculate the line the minimizes the sum of squares of the errors?
]

---

# Bivariate Regression

### Least squares estimation

<br>

$$b = \frac{\sum{(x_i - \bar{x})(y_i - \bar{y})}}{\sum{(x_i - \bar{x})^2}}$$
--

$$a = \bar{y} - b\bar{x}$$
---

# Bivariate Regression

### Least squares estimation


.pull-left[
```{r}
#| label: slope_table
exp_table <- imperf_points1 |> 
  mutate(obs = row_number()) |> 
  dplyr::select(obs, x, y) |> 
  mutate(
    `x - xbar` = x - mean(x), 
    `y - ybar` = y - mean(y), 
    `(x-xbar)(y-ybar)` = `x - xbar` * `y - ybar`, 
    `x - xbar^2` = `x - xbar` ^ 2
  )

x_sum     <- sum(exp_table$x)
y_sum     <- sum(exp_table$y)
x_bar     <- mean(exp_table$x)
y_bar     <- mean(exp_table$y)
xprod_sum <- sum(exp_table$`(x-xbar)(y-ybar)`)
xdev_sum  <- sum(exp_table$`x - xbar^2`)

exp_table |> 
  mutate(obs = as.character(obs)) |> 
  add_row(obs = NA) |> 
  add_row(
    obs = "Sum", 
    x = x_sum, 
    y = y_sum, 
   `(x-xbar)(y-ybar)` = xprod_sum, 
   `x - xbar^2` = xdev_sum) |> 
  add_row(obs = "Mean", x = x_bar, y = y_bar) |> 
  mutate(
    obs = cell_spec(
      x = obs, 
      format = "html", 
      color = if_else(is.na(obs), "white", "black")
    ),
    x = cell_spec(
      x = x, 
      format = "html", 
      color = if_else(is.na(x), "white", "black")
    ),
    y = cell_spec(
      x = y, 
      format = "html", 
      color = if_else(is.na(y), "white", "black")
    ),
    `x - xbar` = cell_spec(
      x = `x - xbar`, 
      format = "html", 
      color = if_else(is.na(`x - xbar`), "white", "black")
    ),
    `y - ybar` = cell_spec(
      x = `y - ybar`, 
      format = "html", 
      color = if_else(is.na(`y - ybar`), "white", "black")
    ),
    `(x-xbar)(y-ybar)` = cell_spec(
      x = `(x-xbar)(y-ybar)`, 
      format = "html", 
      color = if_else(
        condition = is.na(`(x-xbar)(y-ybar)`), 
        true = "white", 
        false = "#cc0033"
      )
    ), 
    `x - xbar^2` = cell_spec(
      x = `x - xbar^2`, 
      format = "html", 
      color = if_else(
        condition = is.na(`x - xbar^2`), 
        true = "white", 
        false = "blue"
      )
    )
  ) |> 
  kable(format = "html", escape = F, align = rep("r", 7))
```
]


.pull-right[

$$b = \frac{\sum{\color{red}{(x_i - \bar{x})(y_i - \bar{y})}}}{\sum{\color{blue}{(x_i - \bar{x})^2}}}$$

$$b = \frac{\color{red}{1}}{\color{blue}{2}}$$

$$a = \bar{y} - b\bar{x}$$

$$a = 2 - (0.5 \times 2)$$

$$a = 1$$

$$\hat{y} = 1 + 0.5x$$
]

---

# Bivariate regression

```{r}
#| label: imperf_plot_bestfit
#| fig-asp: 0.5
imperf_plot_bestfit + 
  annotate(
    geom = "text", x = 1, y = 2.5, size = 7, parse = T, 
    label = TeX("$\\hat{y} = 1 + 0.5x$", output = "character")
  )
```

---

# Bivariate Regression 


- The *best fit line* is determined using the **ordinary least squares** method 

- In other words, we try to find the best line that minimizes 
the .green[distance] between the *predicted values* (the line, $\color{red}{\hat{y}}$) and the observed data, $\color{blue}{y_{i}}$
- The .green[distances] representing deviations from the regression line are called **residuals**.

- The best fit line is the one that reduces the *sum of squares* of the error term (this is a measure of the variability around the best fit line)

- We never measure anything perfectly... there is <u>always</u> error


---

# Fitting a model

$$y_i = \beta_0 + \beta_1 X_i + \epsilon_i$$

- Coefficients (parameter estimates)
  - $\beta_0$: intercept
  - $\beta_1$: slope 
- $\epsilon$: error

We will again use the `mtcars` dataset.

---

# Get to know the dataset

- Using the `mtcars` dataset, we can fit a model with the 
following variables:
  - **response variable**: `mpg`
  - .blue[predictor]: `wt`

```{webr-r}
#| label: webr-mtcars-str
str(mtcars)
```

---

# Fitting the model 

.pull-left[
- We will fit the model using the linear equation...  
$$mpg_i = \beta_0 + \beta_1 wt_i + \epsilon_i$$
- To do this in R, we use the `lm()` function
- `lm()` will use the *ordinary least squares* method to obtain 
parameter estimates, i.e., β<sub>0</sub> (intercept) and 
β<sub>1</sub> (slope)
- In essence, it will estimate the parameters we need to predict 
`mpg` for a given `weight`
]

.pull-right[
```{r}
#| label: mtcars_mod
model <- lm(mpg ~ wt, data = mtcars)
model
```
]

---


<br>
<br>
<br>
<br>
<br>
<br>
<br>

```{r}
#| label: mtcars_p1
#| fig-asp: 0.7
#| fig-align: 'center'
# Basic plot to demonstrate 'y ~ x'
mtcars_p1 <- ggplot(mtcars) + 
  aes(x = wt, y = mpg) + 
  geom_point(
    size = 5, fill = '#cc0033', 
    color = "white", pch = 21, stroke = 1.1
  ) + 
  ds4ling_bw_theme(base_size = 24)
mtcars_p1
```

---

<br>
<br>
<br>
<br>
<br>
<br>
<br>

```{r}
#| label: mtcars_p2
#| fig-asp: 0.7
#| fig-align: 'center'
# Show that mean isnt useful
mtcars_p2 <- mtcars_p1 + 
  geom_hline(
    yintercept = mean(mtcars$mpg), linewidth = 1, 
    color = '#666666'
  ) +
  annotate(
    geom = 'text', 
    fontface = 3, 
    x = (mean(mtcars$wt) + 1), 
    y = (mean(mtcars$mpg) + 1), size = 5,
    label = 'Mean mpg', color = '#666666'
  )
mtcars_p2
```

---

<br>
<br>
<br>
<br>
<br>
<br>
<br>

```{r}
#| label: mtcars_p3
#| fig-asp: 0.7
#| fig-align: 'center'
mod <- lm(mpg ~ wt, data = mtcars)

# show what ordinary least squares means
mtcars_p2 + 
  geom_abline(
    intercept = mod$coef[[1]], 
    slope = mod$coef[[2]], 
    linewidth = 1
  ) + 
  annotate(
    geom = 'text', 
    fontface = 3, 
    x = (mean(mtcars$wt) - 0.5), 
    y = (mean(mtcars$mpg) + 3.5), 
    size = 5, angle = -31, 
    label = 'Best linear fit'
  )
```

---

<br>
<br>
<br>
<br>
<br>
<br>
<br>

```{r}
#| label: mtcars_p4b
#| fig-asp: 0.7
#| fig-align: 'center'

ggplot(mtcars) + 
  aes(x = wt, y = mpg) + 
  geom_hline(
    yintercept = mean(mtcars$mpg), 
    linewidth = 1, color = '#666666'
  ) +
  geom_segment(
    aes(xend = wt, y = fitted(mod), yend = mpg), 
    color = "#666666"
  ) + 
  geom_point(
    size = 5, fill = '#cc0033', 
    color = "white", pch = 21, stroke = 1.1
  ) + 
  stat_smooth(
    method = 'lm', se = F, linewidth = 1, 
    fullrange = TRUE, formula = 'y ~ x', color = "black"
  ) + 
  ds4ling_bw_theme(base_size = 24)
```

---

<br>
<br>
<br>
<br>
<br>
<br>
<br>

```{r}
#| label: mtcars_p4
#| fig-asp: 0.7
#| fig-align: 'center'

ggplot(mod) + 
  aes(x = fitted(mod), y = residuals(mod)) + 
  geom_hline(yintercept = 0, linewidth = 1) + 
  annotate(
    geom = 'text', 
    fontface = 3, 
    x = 10, y = -1, 
    size = 4.5, 
    color = '#666666', 
    label = 'Best linear fit\nLeast squares estimate\n(sideways)'
  ) +
  geom_segment(
    aes(xend = fitted(mod), yend = 0), 
    color = "#666666"
  ) +
  geom_point(
    size = 5, fill = '#cc0033', 
    color = "white", pch = 21, stroke = 1.1
  ) + 
  ds4ling_bw_theme(base_size = 24)
```

---

## Interpreting model output


- Intercept
- Slope
- R<sup>2</sup>

```{r}
#| label: mtcars_table1
mod <- lm(mpg ~ wt, data = mtcars)
print(summary(mod), digits = 2)
```


---

## Interpreting model output

- **Intercept**
- Slope
- R<sup>2</sup>

--

```{r}
#| label: mtcars_p5
#| fig-asp: 0.55
#| fig-align: 'left'
mtcars_p5 <- ggplot(mtcars) + 
  aes(x = wt, y = mpg) + 
  geom_vline(
    xintercept = 0, lty = 2, 
    lwd = 0.8, color = 'black'
  ) + 
  geom_hline(
    yintercept = coef(mod)[1], lty = 2, 
    linewidth = 0.8, color = 'black'
  ) + 
  geom_point(
    size = 5, fill = '#cc0033', 
    color = "white", pch = 21, stroke = 1.1
  ) + 
  stat_smooth(
    method = 'lm', se = F, linewidth = 1, 
    fullrange = TRUE, color = "black", formula = 'y ~ x'
  ) + 
  coord_cartesian(xlim = c(0, 6)) + 
  ds4ling_bw_theme(base_size = 20)
mtcars_p5
```
 

```{r}
#| label: mtcars_table2
mtcars_table2 <- mod |> 
  tidy() |> 
  mutate_if(is.numeric, give_n_digits, n = 2) |> 
  kable(format = "html", align = c("l", rep("r", 4)))
mtcars_table2
```

---

## Interpreting model output

- **Intercept**
- Slope
- R<sup>2</sup>

--

```{r}
#| label: mtcars_p6
#| fig-asp: 0.55
#| fig-align: 'left'
mtcars_p5 + 
  geom_segment(
    aes(x = 1, xend = 0, y = 16, yend = coef(mod)[[1]]), 
    arrow = arrow(angle = 15, type = "closed"), 
    color = 'darkred'
  ) +
  annotate(
    geom = 'text', 
    x = 1.2, y = 15, 
    label = glue("y-intercept = {round(coef(mod)[[1]], 2)}"), 
    size = 5, color = 'darkred'
  )
```

```{r}
#| label: mtcars_table3
mtcars_table2
```

---

## Interpreting model output

- Intercept
- **Slope**
- R<sup>2</sup>

--

```{r}
#| label: mtcars_p7
#| fig-asp: 0.55
#| fig-align: 'left'
mtcars_p5 + 
  geom_segment(
    aes(x = 0, xend = 1, y = coef(mod)[1], yend = coef(mod)[1]), 
    lty = 2, linewidth = 0.8, color = 'black'
  ) + 
  geom_segment(
    mapping = aes(
      x = 1, xend = 1, 
      y = coef(mod)[1], yend = coef(mod)[1] + coef(mod)[2]
    ), 
    lty = 2, linewidth = 0.8, color = 'black', 
  ) + 
  geom_segment(
    mapping = aes(x = 2, xend = 1.1, y = (coef(mod)[1]) - 1), 
    yend = coef(mod)[1], 
    arrow = arrow(angle = 15, type = "closed"), 
    color = 'darkred'
  ) +
  annotate(
    geom = 'text', 
    x = 2.5, y = (coef(mod)[1] - 1), 
    label = glue("Slope = {round(coef(mod)[[2]], 2)}"), 
    size = 5, color = 'darkred'
  )
```

```{r}
#| label: mtcars_table4
mtcars_table2
```


---

## Interpreting model output

- Intercept
- Slope
- **R<sup>2</sup>**

--

The coefficient of determination

- An overall assessment of model fit
- R<sup>2</sup> is the variance explained by your model
- Ranges from 0 to 1 (1 = 100% variance explained)

---

## Making predictions

Recall the linear model equation...

$$\hat{y} = \beta_0 + \beta_1 wt_i + \epsilon_i$$

- Our `mtcars` model can be summarized as...

```{r}
#| label: model_print
#| results: 'asis'
b <- coef(mod)
cat(sprintf("$$mpg = %.02f + %.02f wt$$", b[1], b[2]))
```


- What is the predicted `mpg` for a car that weighs 1 unit? And one that weighs 3? And 6?

--

`r round(coef(mod)[2] * 1 + coef(mod)[1], 2)` $mpg = 37.29 + -5.34 \times 1$

`r round(coef(mod)[2] * 3 + coef(mod)[1], 2)` $mpg = 37.29 + -5.34 \times 3$ 

`r round(coef(mod)[2] * 6 + coef(mod)[1], 2)` $mpg = 37.29 + -5.34 \times 6$


---

<br>
<br>
<br>
<br>
<br>
<br>
<br>
```{r}
#| label: mtcars_p8
#| fig-asp: 0.7

ggplot(mtcars) + 
  aes(x = wt, y = mpg) + 
  geom_smooth(
    method = 'lm', se = F, 
    color = 'white', formula = 'y ~ x'
  ) + 
  geom_point(size = 4) + 
  geom_point(size = 3, color = 'lightblue') + 
  stat_smooth(method = "lm", fullrange = TRUE, formula = 'y ~ x') + 
  geom_segment(
    mapping = aes(
      x = 0, xend = 1, 
      y = round(coef(mod)[2] * 1 + coef(mod)[1], 2), 
      yend = round(coef(mod)[2] * 1 + coef(mod)[1], 2)
    ), 
    lwd = 1, lty = 3, color = "darkred"
  ) + 
  geom_segment(
    mapping = aes(
      x = 0, xend = 3, 
      y = round(coef(mod)[2] * 3 + coef(mod)[1], 2), 
      yend = round(coef(mod)[2] * 3 + coef(mod)[1], 2)
    ), 
    lwd = 1, lty = 3, color = "darkred"
  ) + 
  geom_segment(
    mapping = aes(
      x = 0, xend = 6, 
      y = round(coef(mod)[2] * 6 + coef(mod)[1], 2), 
      yend = round(coef(mod)[2] * 6 + coef(mod)[1], 2)
    ), 
    lwd = 1, lty = 3, color = "darkred"
  ) +
  scale_x_continuous(limits=c(0, 6.25), expand = c(0, 0)) +
  ds4ling_bw_theme(base_size = 24)
```