---
title: "Week 6: Multiple Regression"
subtitle: 'Statistics for Linguists'
author  : "Kyle Parrish"
date    : "Goethe Uni Frankurt</br>Summer 2025</br>Last update: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, hygge, my-css.css, rutgers-fonts]
    lib_dir: libs
    nature:
      countIncrementalSlides: false
      ratio: 16:9
---

```{r, include = FALSE}
# Packages --------------------------------------------------------------------

# Custom
library("ds4ling")
library("lingStuff")
library("untidydata")
library("academicWriteR")

# Tidyverse
library("dplyr")
library("tidyr")
library("readr")
library("purrr")
library("here")
library("forcats")
library("stringr")
library("glue")

# Modeling
library("lme4")
library("brms")

# Plotting
library("ggplot2")
library("patchwork")
library("ggridges")
library("ggrepel")
library("viridis")
library("ggfortify")

# Misc
library("knitr")
library("kableExtra")
library("latex2exp")
library("broom")
library("fGarch")
library("equatiomatic")
library("countdown")
library("tiktokrmd")
```


```{r xaringanExtra-clipboard, echo=FALSE}
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r}
#| echo: false
source("_common.R")
```


<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
.center[
## Slides are largely adapted from those by Joseph Casillas [here](https://www.ds4ling.jvcasillas.com/slides/05_lm/03_mrc/#/title-slide)]

---

# Linear Regression

```{r include=FALSE}
knitr::opts_chunk$set(cache=F, warning=FALSE, message=FALSE, echo = FALSE)
library(arm)
library(ggplot2)
library(tidyr)
library(dplyr)
options(scipen = 0)
```

.pull-left[
.large[
So far, we have been focusing on **comparing two means** (e.g. is class A better than class B at math).
]
]

--

.pull-right[
.large[
Last week we did a .green[bivariate linear regression] (how much does fuel efficiency decrease as weight of a car increases?)

We also mentioned we would come back to .red[confounding variables]
]]

---

